Prestige bias in cultural evolutionary dynamics

### 2023-02-06 ms/ms.tex
reduce words to 9.5k

\begin{abstract}
A common bias in cultural transmission is to copy whoever appears more successful. This success bias only depends on the performance of the role-model.
Because success may be difficult to estimate, it has been suggested that individuals may indirectly estimate success by using prestige as a proxy for success. However, prestige can be defined in different ways: it can reflect performance in another, possibly related, task; but it can also reflect overall influence of the role-model, irrespective of performance.
Here, we model prestige bias as a combination of indirect success bias and influence bias, the latter meaning that the choice of a role-model depends on the number of individuals that have already copied from that role-model.
We analyze the effect of prestige on cultural evolutionary dynamics using mathematical analysis and stochastic simulations.
We find analytic approximations to the stochastic role-model choice process, which facilitate further mathematical analysis and reduce the computational complexity of simulations. We validate these approximations and demonstrate their robustness to model assumptions.
We also find approximations to the fixation probability and the fixation time of an invading cultural trait, in both constant and changing environments. 
These approximations are similar to Kimura's approximations for population-genetic models.
These similarities show that indirect success bias effectively plays the role of natural selection, whereas influence bias effectively plays the role of genetic drift.	
Influence bias also accelerates the evolutionary dynamics, as can be expected in a \emph{rich-getting-richer} process.
Our model extends the literature that describes and analyses cultural transmission, 
especially in human societies where social media is popular and thus influence bias may be expected. 
Further work is needed to test if this model could explain and predict various phenomena in cultural evolution.
\end{abstract}

@@

\paragraph{Empirical evidence of prestige bias}
\citet{prestige_cultural_learning} report the first direct tests in children that suggest the existence of prestige bias, defined as the tendency to learn from individuals to whom others have preferentially attended, learned, or deferred.
Their definition of prestige is similar to our influence bias. They showed that the odds of 3-4 years-old children learning from an adult role-model to whom bystanders had previously preferentially attended for 10 seconds were more than twice those of their learning from a role-models whom bystanders ignored.
They also note that prestige effects are domain sensitive: they found that prestigious role-models were attended more when demonstrating artifact use, whereas role-models presenting food preferences had less attendants, suggesting that the domain itself (artifact use vs. food preference) can affect the attendance, and hence the prestige of the role-model.
This lead \citet{prestige_cultural_learning} to suggest that when the trait is costly to learn individually, prestige will have a stronger bias.
It would be interesting to include costs in our model to try and observe these effects and dynamics in a large population.

@@

\paragraph{Conclusions.}
Here, we studied a model of cultural evolution under two transmission biases: the commonly studied success bias, together with influence bias, which has so far received less attention. We found approximations for this complex dynamics. We then showed that success bias affects the evolutionary dynamics much like natural selection does, whereas influence bias has a similar effect to random genetic drift. We also find a clear advantage to individuals that can choose the relative weight of the two biases.

>>>>

\begin{abstract}
A common bias in cultural transmission is to copy whoever appears more successful. This success bias only depends on the performance of the role-model.
Because success may be difficult to estimate, it has been suggested that individuals may indirectly estimate success by using prestige as a proxy for success. Here, we model prestige bias as a combination of indirect success bias and influence bias, the latter meaning that the choice of a role-model depends on the number of individuals that have already copied from that role-model.
We analyze the effect of prestige on cultural evolutionary dynamics using mathematical analysis and stochastic simulations.
We find analytic approximations to the stochastic role-model choice process, which facilitate further mathematical analysis and reduce the computational complexity of simulations. We also find approximations to the fixation probability and the fixation time of an invading cultural trait in different environments. 
We show that indirect success bias effectively plays the role of natural selection, whereas influence bias effectively plays the role of genetic drift.	
Influence bias also accelerates the evolutionary dynamics, as can be expected in a \emph{rich-getting-richer} process.
Our model describes and analyses cultural transmission, especially in human societies where social media is popular.

\end{abstract}

@@

\iffalse
\paragraph{Empirical evidence of prestige bias}
\citet{prestige_cultural_learning} report the first direct tests in children that suggest the existence of prestige bias, defined as the tendency to learn from individuals to whom others have preferentially attended, learned, or deferred.
Their definition of prestige is similar to our influence bias. They showed that the odds of 3-4 years-old children learning from an adult role-model to whom bystanders had previously preferentially attended for 10 seconds were more than twice those of their learning from a role-models whom bystanders ignored.
They also note that prestige effects are domain sensitive: they found that prestigious role-models were attended more when demonstrating artifact use, whereas role-models presenting food preferences had less attendants, suggesting that the domain itself (artifact use vs. food preference) can affect the attendance, and hence the prestige of the role-model.
This lead \citet{prestige_cultural_learning} to suggest that when the trait is costly to learn individually, prestige will have a stronger bias.
It would be interesting to include costs in our model to try and observe these effects and dynamics in a large population.

@@

\fi
\paragraph{Conclusions.}
Here, we studied a model of cultural evolution under two transmission biases: the commonly studied success bias, together with influence bias, which has so far received less attention. We found approximations for this complex dynamics. We then showed that success bias affects the evolutionary dynamics much like natural selection does, whereas influence bias has a similar effect to random genetic drift. We also find a clear advantage to individuals that can choose the relative weight of the two biases.
